Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-01 paper_jasa.tex
Organized section 5, Simulation experiments. A lot more needs to be added to this section.

However, measuring the strength of patterns seen in plots, and differences in individual perceptual ability, is something that is difficult and perhaps handicaps graphics  use among statisticians, where measuring probabilities is of primary importance. This has also been addressed in recent research. \citet{buja:2009} proposes a protocol that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis. One of the protocols, the lineup, places the data plot among  a page of plots of null data, and asks a human judge to pick the plot that is different. Figure \ref{fig:test_category} shows an example lineup. Which plot do you think is the most different from the others? Wrapped in a process that mirrors classical inference, where there is an explicit, a priori, null hypothesis, picking the plot of the data from the null plots represents a rejection of that null hypothesis. The null hypothesis typically derives from the task at hand, or the type of plot being made. The alternative encompasses all possible antitheses, all types of patterns that might be detected in the data plot, accounting for all possible deviations from the null without the requirement to specify these ahead of time. The probability of rejection can be quantified, along with Type I, and Type II error, and $p$-value and power can be defined and estimated.

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{test_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the the largest shift or location difference between the boxplots?}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

%\red{How are $\Theta_0$ and $\Theta^c_0$ defined? - why do we need them? wouldn't it be better to re-formulate power in terms of acceptance region, not in terms of the parameter space? - $1/m$ also doesn't look right for the power - I think that's just looking at one observer.}
%\begin{dfn} \label{dfn:power}
%The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
%    \begin{equation*}
%      \text{Power}_V(\theta)= 
%        \begin{cases} 
%              \frac1m & \text{if $\theta \in \Theta_0$,} \\
%              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
%        \end{cases}
%    \end{equation*}
%\end{dfn}
%
%
%\red{Alternative Power definition:}
\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= Pr(\text{Reject } H_0) = P(X \ge x_{\alpha}) = 
        \begin{cases} 
               1 - \text{Binom}_{K, 1/m}(x_{\alpha} - 1) & \text{ if } H_0 \text{ is true }.\\
              1 - F_{X, \theta} (x_{\alpha} - 1) & \text{ if } H_1 \text{ is true}.
        \end{cases}
    \end{equation*}
\end{dfn}

@@

\section{Application of Visual Inference in Linear Models} \label{sec:regression}

@@

\section{Results of Simulation Experiments} \label{sec:simulation}

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression scenario. There are three experiments. The first two are classical settings where we would not expect the lineup protocol to do better than the traditional test, but we hope it to perform favorably. The third experiment is a scenario where assumptions required for the traditional test are violated, and we would expect the lineup protocol to outperform the traditional test.

\subsection{Discrete covariate}

The experiment is designed to study the ability of human observers to detect the effect of a single variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \eqref{multi}). Data is simulated for different values of $\beta_2 (=0, 1, 3, 5, 7, 8, 10, 16)$, with two sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The set of $\beta_2$ values was chosen so that estimates of the power should produce reasonable power curves, comparable to the theoretical UMP test. We fixed the values of $\beta_0 = 5$,  $\beta_1=15$ and values for $X_1$ were generated as a random sample from a Poisson(30) distribution. Data sets with different combinations of $\beta_2$,  $n$ and $\sigma$ were generated with frequencies shown in Table \ref{tbl:experiment_params}. Three replicates of each level were generated. These produced 60 different ``actual data sets''.

\begin{table}[hbtp]
\caption{Values of parameters considered for three survey experiments} % title name of the table
\centering
% centering table
\begin{tabular}{c c l l l}
% creating 10 columns
\hline\hline

% inserting double-line
& & \multicolumn{3}{c}{values for slope ($\beta$)} \\
 \cline{3-5}

\raisebox{1.5ex}  {Sample size ($n$)} &   \raisebox{1.5ex} {$\sigma$} &  \multicolumn{1}{c} {Experiment 1}  & \multicolumn{1}{c} {Experiment 2}  & \multicolumn{1}{c} {Experiment 3}
\\ [0.5ex]
\hline
% inserts single-line

@@

For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the traditional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

This experiment is designed for the same regression model (Equation \eqref{multi}) with $p=1$ and covariate $X_1$ being continuous. Data is simulated with two sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The different values for slope parameter that are used to generate data are shown in Table \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ while generating actual data. For each combination of parameter, we generate at least three different actual data sets constituting a total of 65 lineups.

@@

@@

@@

>>>>

However, measuring the strength of patterns seen in plots, and differences in individual perceptual ability, is something that is difficult and perhaps handicaps graphics  use among statisticians, where measuring probabilities is of primary importance. This has also been addressed in recent research. \citet{buja:2009} proposes a protocol that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis. One of the protocols, the lineup, places the data plot among  a page of plots of null data, and asks a human judge to pick the plot that is different. Figure \ref{fig:test_category} shows an example lineup. Which plot do you think is the most different from the others? (The position of the actual data plot is provided in Section \ref{sec:category}.) Wrapped in a process that mirrors classical inference, where there is an explicit, a priori, null hypothesis, picking the plot of the data from the null plots represents a rejection of that null hypothesis. The null hypothesis typically derives from the task at hand, or the type of plot being made. The alternative encompasses all possible antitheses, all types of patterns that might be detected in the data plot, accounting for all possible deviations from the null without the requirement to specify these ahead of time. The probability of rejection can be quantified, along with Type I, and Type II error, and $p$-value and power can be defined and estimated.

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{test_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

%\red{How are $\Theta_0$ and $\Theta^c_0$ defined? - why do we need them? wouldn't it be better to re-formulate power in terms of acceptance region, not in terms of the parameter space? - $1/m$ also doesn't look right for the power - I think that's just looking at one observer.}
%\begin{dfn} \label{dfn:power}
%The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
%    \begin{equation*}
%      \text{Power}_V(\theta)= 
%        \begin{cases} 
%              \frac1m & \text{if $\theta \in \Theta_0$,} \\
%              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
%        \end{cases}
%    \end{equation*}
%\end{dfn}
%
%
%\red{Alternative Power definition:}
\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= Pr(\text{Reject } H_0) = P(X \ge x_{\alpha}) = 
        \begin{cases} 
               1 - \text{Binom}_{K, 1/m}(x_{\alpha} - 1) & \text{ if } H_0 \text{ is true}.\\
              1 - F_{X, \theta} (x_{\alpha} - 1) & \text{ if } H_1 \text{ is true}.
        \end{cases}
    \end{equation*}
\end{dfn}

@@

\section{Application to Linear Models} \label{sec:regression}

@@

\section{Simulation Experiments} \label{sec:simulation}

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression scenario. There are three experiments. The first two are classical settings where we would not expect the lineup protocol to do better than the traditional test, but we hope it to perform favorably. The third experiment is a scenario where assumptions required for the traditional test are violated, and we would expect the lineup protocol to outperform the traditional test. (Data and lineups used in the experiments are available in the supplementary material.)

\subsection{Discrete covariate}\label{sec:category}

The experiment is designed to study the ability of human observers to detect the effect of a single variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical UMP test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson(30) distribution. \green{Explain why this was used.} Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets''.

Each subject viewed 10 lineups. Selection of lineups shown to each subject followed these rules .... \green{fill this in}. Figure \ref{fig:test_category} is an example of one of these lineups. The visual test statistic is a side-by-side boxplot.

\green{Did you in fact bin like this???} For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the traditional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

\begin{table}[hbtp]
\caption{Combination of parameter values, $\beta_2$,  $n$ and $\sigma$, used for each of the simulation experiments.} % title name of the table
\centering
% centering table
\begin{tabular}{c c c c c}
% creating 10 columns
\hline\hline

% inserting double-line
& & \multicolumn{3}{c}{Slope ($\beta$)} \\
 \cline{3-5}

Sample size  & Error SD  &  \multicolumn{1}{c} {Experiment 1}  & \multicolumn{1}{c} {Experiment 2}  & \multicolumn{1}{c} {Experiment 3} \\
 ($n$) &   ($\sigma$) & Discrete covariate & Continuous covariate & Contaminated data 
\\ [0.5ex]
\hline
% inserts single-line

@@

@@

This experiment is designed for the same regression model (Equation \ref{multi}) with $p=1$ and covariate $X_1$ being continuous. Data is simulated with two sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The different values for slope parameter that are used to generate data are shown in Table \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ while generating actual data. For each combination of parameter, we generate at least three different actual data sets constituting a total of 65 lineups.

\green{Fill in details of lineups, probably need a figure with an example lineup, and how lineups were assigned to subjects.}

@@

\green{Fill in details of lineups, probably need a figure with an example lineup, and how lineups were assigned to subjects. Show a sample lineup, probably the most difficult, instead of the single plot.}

@@

\subsection{Results}

@@

\green{A lot more needs to be done here!!!}
